Investigating the applications and limits of single particle light scattering using limited detection schemes

### 2022-11-07 Manuscript.tex
Add files via upload

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 190 2020-11-23 11:12:32Z rishi $
%%
%%
\documentclass[review,3p]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage{babel}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

@@

\author[aff1]{Dan Maciver}
\affiliation[aff1]{organization={Department of Chemical Engineering,
			University of Strathclyde},
            addressline={16 Richmond Street}, 
            city={Glasgow},
            postcode={G1 1XQ}, 
            country={Scotland}}
\author[aff1]{Praveen Parthasarathi}
\author[aff1]{Mark Haw}
\author[aff1]{Leo Lue}
\author[aff1]{Jan Sefcik}

@@

\begin{keyword}
	Optical Trapping \sep Light Scattering \sep Measurements \sep Statistics

@@

Consider a dimer with unequal sphere diameters $a_1, a_2$ in an optical trap positioned in some orientation $\hat{s}$. Due to the Brownian motion from the surrounding fluid the dimer's orientation changes with time. Assuming we cannot accurately determine the dimer's orientation from imaging alone we instead rely on it's light scattering. We collect the near field intensity, via optical fibres,from a probe laser at 3 predetermined angles $\theta_1, \theta_2, \theta_3$. To determine the particle's orientation we compare the produced light scattering signal to a collection of reference orientations that are evenly spaced around the orientation space.

@@

	\begin{eqnarray}
		p(y(\hat{s})\parallel\hat{n})&= \Pi^3_{k=1} 
		(2\pi\sigma_k^2)^{-1/2} e^{-(y(\hat{n})_k-y(\hat{s})_k)^2/2\sigma_k^2}
	\end{eqnarray}

Where $\sigma_k$ is the average error in our signal collection; we left this as a tuneable variable to measure how the influence of experimental error would effect our predictive model. This was implemented by adjusting the values of $y(\hat{s}_k)$. 
\begin{eqnarray}
	\sigma_k = 0.17\bar{y}(\hat{n}_k) \\
	\Rightarrow y(\hat{s}_k) = y(\hat{s})_k \pm n\sigma_k \\ 
	n \in \mathbb{R}, n\in[0,1]
\end{eqnarray}

@@

\begin{eqnarray}
	p(\hat{n}\parallel y(\hat{s}))&= \frac{p(y(\hat{s})\parallel\hat{n})p(\hat{n})}{p(y)}
\end{eqnarray}

@@

\begin{eqnarray}
	p(y) = \int p(y(\hat{s})\parallel \hat{n}) p(\hat{n}) d\hat{n}
\end{eqnarray}
In order to test the accuracy of our model we first test it using an idealised simulation of the target dimer.

\subsection{Testing the model}
\label{2.1}
We use the Brownian OT package by Jerome Fung \cite{Brownian_OT} to simulate how our target dimer moves while subject to an optical trap. At each time step we determine the dimer's orientation and use MSTM to calculate its scattering pattern. We then use the above method to estimate the dimer orientation based on its scattering pattern. The priori estimate $p(\hat{n})$ can be defined as a Boltzmann's distribution based on the dot product of our previous estimation $\hat{n}_{t-\Delta t}$ and each reference orientation $\hat{n}_t$:
\begin{eqnarray}
	r_j(t, \hat{n})= \hat{n}_{t} \cdot \hat{n}_{t-\Delta t} \\
	p(\hat{n})= \frac{e^{(r_j(t,\hat{n})-1)}}
	{\Sigma_{i=1}^{24}e^{r_i(t, \hat{n})}}
\end{eqnarray}

Since we are testing with a simulation we already know the exact orientation of the dimer, we want to know how often our model can determine the closest reference orientation $\hat{n}$ to the true orientation $\hat{s}$, and how precise our prediction is. For a perfect model our probability distribution should equal 1 at the correct result and zero elsewhere. In reality our model will be some probability distribution that has some probability of any orientation being correct. To evaluate our model's accuracy and confidence we calculate the Kullback-Leibler divergence of the two probability distributions.

\begin{eqnarray}
	K_{l, \#}(p_{guess} \parallel p_{ideal}) &= 
	p_{ideal} \ln \left[
	\frac{p_{ideal}}{p(\hat{n}\parallel y_k(\hat{s}))}
	\right]
\end{eqnarray}

@@

\begin{eqnarray}
	K_{l,total} = \sum\limits_{\# =1}^{timesteps} K_{l,\#} \\
	F(K_l) = \frac{K_{l,\ worst}}{K_{l,total}}
\end{eqnarray}

@@

\subparagraph{Optimization of angle choice}
\label{2.2}
The efficacy of our estimation is a function of our detection angles, testing every possible configuration manually is time consuming, therefore we utilised Ultra nest to apply a Monte Carlo simulation to our model. Ultra-nest is a sophisticated python package for fitting models with complex parameter interactions \cite{Ultranest}. Ultra-nest implements a variation of Monte Carlo integration,nested-sampling, where the likelihood contour is used to update a group of live points chosen from the prior estimation.

We used the Ultra-nest software to create a set of live points of different choices of $\theta_1, \theta_2, \ \& \ \theta_3$, with each point being evaluated by the divergence ($K_{l, total}$) produced. After evaluating every live point, the worst results are removed and new live points are created based on the information provided by the previous live points. This results in a convergence on parameter choices that result in an average divergence result of almost 0. A common issue with complex parameter spaces, such as ours, is the issue of degeneracies (where two or more parameters have identical parameter spaces), to combat this we implemented a random step sampler that selects a new live point some 'n' steps away in the parameter space.

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width =\textwidth]{traj.png}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{pos.png}
	\end{subfigure}
	\caption{Simulation results of: a) the dimers orientation vector with time, b) the dimer's [x,y,z] position with time.}
\end{figure}

@@

We first tested how varying the error value would effect our model's efficacy; calculating the improvement factor $F(K_l)$ with each test. As can be seen from \figurename{ 3} our model actually performs better with noisy signals, to the extent that a signal error of $10\%$ resulted in a worse performance than if we had just guessed each orientation at random. 
\begin{figure}
	\begin{subfigure}{0.33\textwidth}
		\includegraphics[width=\textwidth]{Error_10_percent.png}
		\subcaption{$F(K_l) = 0.9422$}
	\end{subfigure}
	\begin{subfigure}{0.33\textwidth}
		\includegraphics[width=\textwidth]{Error_20_percent.png}
		\subcaption{$F(K_l) = 1.366$}
	\end{subfigure}
	\begin{subfigure}{0.33\textwidth}
		\includegraphics[width=\textwidth]{Error_30_percent.png}
		\subcaption{$F(K_l)=1.440$}
	\end{subfigure}
	\caption{Model predictions with signal error at a) $10 \%$, b)  $20\%$, and c) $30\%$}
\end{figure}

The issue with having a high signal error is that our model's prediction can be very far off from the correct answer and still be confident with it's results; fortunately our priori for the dimer's orientation can help to correct our estimate by limiting the probability in the physical space. We found that the inclusion of our priori estimate compared to a uniform priori improved our results by around $20 \%$. We can alter the influence of our priori by adding an additional exponential

\subsection{Ultra-nest results}
\label{3.3}
We first allowed ultra nest to traverse the parameter space while each angle was constrained to a predefined area. We assumed that the each fibre would be placed within a roughly $30^{\circ}$ range so that we get a sampling of the entire light scattering pattern, we didn't search further than $100^{\circ}$ as the signals would be to low to detect for a micron scale scatterer. The sampler ran until the divergence result was in the order of $10^1$.

\begin{figure} [t]
	\centering
	\includegraphics[width=0.4\textwidth]{Cornerangleconstrained-1.png}
	\caption{Corner plot from ultra-nest, darker regions indicate a better divergence result. $\theta_1$ was constrained between $15^{\circ}$ and $25^{\circ}$; $\theta_2$ was constrained between $30^{\circ}$ and $60^{\circ}$; $\theta_3$ was constrained between $65^{\circ}$ and $100^{\circ}$.}
\end{figure}
As can be seen from \figurename{ 4} the parameter space for our angle choice is essentially flat, with no explicit maxima to be found, the sudden increase in our divergence around $\theta_1 = 18^{\circ}$ is likely due to the fact that if we consider the scattering curves for our reference orientations there exist local minima on at roughly $\theta_1 = 13^{\circ}$ and $25^{\circ}$. Therefore between these two minima the model struggles to differentiate between reference orientations given that the light intensity is similar for different orientations.

>>>>

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 190 2020-11-23 11:12:32Z rishi $
%%
%%
\documentclass[final,3p, twocolumn]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb, amsthm}
\usepackage{amsmath}
%% The amsthm package provides extended theorem environment
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage{babel}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

@@

\author[aff1]{Dan Maciver\corref{cor1}} %
\ead{Daniel.Maciver.2016@uni.strath.ac.uk}

\author[aff1]{Praveen Parthasarathi}

\author[aff1]{Mark Haw}

\author[aff1]{Leo Lue}

\author[aff1]{Jan Sefcik}

\cortext[cor1]{Corresponding author}
\affiliation[aff1]{organization={Department of Chemical Engineering,
			University of Strathclyde},
            addressline={16 Richmond Street}, 
            city={Glasgow},
            postcode={G1 1XQ}, 
            country={Scotland}}

@@

\begin{keyword}
	Optical Trapping \sep Light Scattering \sep Measurements \sep Bayesian Statistics

@@

Consider a dimer with unequal sphere diameters $a_1, a_2$ in an optical trap positioned in some orientation $\hat{s}$. Due to the Brownian motion from the surrounding fluid the dimer's orientation changes with time. Assuming we cannot accurately determine the dimer's orientation from imaging alone we instead rely on it's light scattering. We collect the near field intensity, via optical fibres,from a probe laser at 3 angles $\theta_1, \theta_2, \theta_3$. To classify the particle's orientation we compare the produced light scattering signal to a collection of reference orientations that are evenly spaced around the orientation space.

@@

\begin{align}
	p(y(\hat{s})\parallel\hat{n}) =& \Pi^3_{k=1}
	(2\pi\sigma_k^2)^{-1/2} \nonumber \\     &e^{-(y(\hat{n})_k-y(\hat{s})_k)^2/2\sigma_k^2}
\end{align}

Where $\sigma_k$ is the average error in our signal collection; we left this as a tuneable variable to measure how the influence of signal error would effect our predictive model. This was implemented by adjusting the values of $y(\hat{s}_k)$.

\begin{align}
	\centering
	\sigma_k =& 0.17\bar{y}(\hat{n}_k) \\
	\Rightarrow y(\hat{s}_k) =& y(\hat{s})_k \pm n\sigma_k \\ 
	n \in \mathbb{R}, \ \ & n\in[0,1] \nonumber
\end{align}

@@

\begin{align}
	p(\hat{n}\parallel y(\hat{s}))&= \frac{p(y(\hat{s})\parallel\hat{n})p(\hat{n})}{p(y)}
\end{align}

@@

\begin{align}
	p(y) = \int p(y(\hat{s})\parallel \hat{n}) p(\hat{n}) d\hat{n}
\end{align}

The inclusion of $p(y)$ normalises our probability distribution. In order to test the accuracy of our model we first test it using an idealised simulation of the target dimer.

\subsection{Testing the model}
\label{2.1}
We use the Brownian OT package by Jerome Fung \cite{Brownian_OT} to simulate how our target dimer moves while subject to an optical trap. At each time step we determine the dimer's orientation and use MSTM to calculate its scattering pattern. We then use the above method to estimate the dimer orientation based on its scattering pattern. The priori estimate $p(\hat{n})$ can be defined as a Boltzmann's distribution based on the normalised dot product of our previous estimation $\hat{n}_{t-\Delta t}$ and each reference orientation $\hat{n}_t$:

\begin{align}
	r_j(t, \hat{n})= \frac{1}{2}((\hat{n}_{t} \cdot \hat{n}_{t-\Delta t})-1) \\
	p(\hat{n})= \frac{e^{\beta (r_j(t,\hat{n})-1)}}
	{\Sigma_{i=1}^{24}e^{\beta(r_i(t, \hat{n}))}}
\end{align}

Where $\beta$ is a weighting factor that controls the significance of our priori estimate. Since we are testing with a simulation we already know the exact orientation of the dimer, we want to know how often our model can determine the closest reference orientation $\hat{n}$ to the true orientation $\hat{s}$, and how precise our prediction is. For a perfect model our probability distribution should equal 1 at the correct result and zero elsewhere. In reality our model will be some probability distribution that has some probability of any orientation being correct. To evaluate our model's accuracy and confidence we calculate the Kullback-Leibler divergence of the two probability distributions.

\begin{align}
	K_{l, \#}(p_{guess} \parallel p_{ideal}) &= 
	p_{ideal} \nonumber \\ 
	&\ln \left[\frac{p_{ideal}}{p(\hat{n}\parallel y_k(\hat{s}))}
	\right]
\end{align}

@@

\begin{align}
	K_{l,total} = \sum\limits_{\# =1}^{timesteps} K_{l,\#} \\
	F(K_l) = \frac{K_{l,\ worst}}{K_{l,total}}
\end{align}

@@

\subparagraph{Optimization of parameters}
\label{2.2}
The efficacy of our estimation is a function of our detection angles, testing every possible configuration manually is time consuming, therefore we utilised Ultra nest to apply a Monte Carlo simulation to our model. Ultra-nest is a sophisticated python package for fitting models with complex parameter interactions \cite{Ultranest}. Ultra-nest implements a variation of Monte Carlo integration,nested-sampling, where the likelihood contour is used to update a group of live points chosen from the prior estimation ($0^{\circ} \leq \theta_k \leq 180^{\circ}$).

We used the Ultra-nest software to create a set of live points of different choices of $\theta_1, \ \theta_2, \ \& \ \theta_3$, with each point being evaluated by its improvement factor ($F(K_l)$) produced. After evaluating every live point, the worst results are removed and new live points are created based on the information provided by the previous live points. This results in a convergence on parameter choices that result in an average divergence result of almost 0. A common issue with complex parameter spaces, such as ours, is the issue of degeneracies (where two or more parameters have identical parameter spaces), to combat this we implemented a random step sampler that selects a new live point some 'n' steps away in the parameter space.

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width =\textwidth]{traj.png}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{pos.png}
	\end{subfigure}
	\caption{Simulation results of: a) the dimers orientation vector with time, b) the dimer's [x,y,z] position with time.}
\end{figure}

@@

We first tested how varying the error value would effect our model's efficacy; calculating the improvement factor $F(K_l)$ with each test, the measurements were made with detection angles of $15^{\circ}, \ 45^{\circ}, \ \& \ 90^{\circ}$. As can be seen from \figurename{ 3} our model actually performs better with noisy signals, to the extent that there was a 0.2 difference between our estimation with $10 \%$ signal error and $30\%$. The reasoning for this being that a noisy signal means that any time our model deviates from the correct response it will quickly self correct due to the larger variance.  The issue with having a high signal error is that our model's prediction can be very far off from the correct answer and still be confident with it's results; fortunately our priori for the dimer's orientation can help to correct our estimate by limiting the probability in the physical space. We found that the inclusion of our priori estimate compared to a uniform priori improved our results by around $20 \%$.

\begin{figure}[t]
	\centering
	\begin{subfigure}{0.33\textwidth}
		\includegraphics[width=\textwidth]{Error_10_percent.png}
		\subcaption{$F(K_l) = 0.9422$}
	\end{subfigure}
	\begin{subfigure}{0.33\textwidth}
		\includegraphics[width=\textwidth]{Error_20_percent.png}
		\subcaption{$F(K_l) = 1.366$}
	\end{subfigure}
	\begin{subfigure}{0.33\textwidth}
		\includegraphics[width=\textwidth]{Error_30_percent.png}
		\subcaption{$F(K_l)=1.440$}
	\end{subfigure}
	\caption{Model predictions with signal error at a) $10 \%$, b)  $20\%$, and c) $30\%$. Dotted line represents minimum distance between reference orientations}
\end{figure}

We estimate that at minimum our experimental setup should have a $17\%$ signal error, assuming near perfect alignment of fibres and minimal interference from other microspheres in the solution. Therefore the fact that our model provides a better estimation of the dimer's motion with worse noise is promising for the implementation of our model into a experimental setup. To assess the impact of signal error, angle choice, and weighting factor we utilised ultranest to step through the entire parameter space.

\subsection{Ultra-nest results}
\label{3.3}
We first wanted to measure analyse the impact of our choice of angles in our predictive model.

\begin{figure} [t]
	\centering
	\includegraphics[width=0.4\textwidth]{Cornerangleconstrained-1.png}
	\caption{Corner plot from ultra-nest, darker regions indicate a better divergence result. $\theta_1$ was constrained between $15^{\circ}$ and $25^{\circ}$; $\theta_2$ was constrained between $30^{\circ}$ and $60^{\circ}$; $\theta_3$ was constrained between $65^{\circ}$ and $100^{\circ}$.}
\end{figure}

As can be seen from \figurename{ 4} the parameter space for our angle choice is essentially flat, with no explicit maxima to be found, the sudden increase in our divergence around $\theta_1 = 18^{\circ}$ is likely due to the fact that if we consider the scattering curves for our reference orientations there exist local minima on at roughly $\theta_1 = 13^{\circ}$ and $25^{\circ}$. Therefore between these two minima the model struggles to differentiate between reference orientations given that the light intensity is similar for different orientations.
